Leadership across cultures and contexts: an exploratory analysis

### 2020-05-14 leadership_across_cultures_contexts.Rmd
Starting to revise abstract.

Leader-follower dynamics are central to much of social life. Processes of leadership and followership observed in traditional societies, especially among small-scale societies, are often quite distinct from those observed among WEIRD societies and their complex organizations. Many researchers have turned to evolutionary theory to integrate diverse perspectives on human leadership and to better understand the diversity of leadership and followership observed across human cultures and across social contexts. Evolutionary theories of leadership, in turn, draw on ethnographic case studies from traditional societies that are thought to more closely resemble the small, face-to-face societies in which humans evolved. Currently, though, there is limited systematic data on the nature of leadership in such societies.

We build on our previous cross-cultural findings to systematically review the nature of leadership across the full range of human cultural diversity. Over 1200 ethnographic texts from the Human Relations Area Files (HRAF) were coded on leader qualities, their functions, and the costs and benefits for both leaders and followers across diverse social contexts. We discovered evidence for both cultural universals in leadership, as well as important variation, especially when compared to leadership in WEIRD societies. Across all contexts, leaders in the ethnographic record are most commonly described as intelligent and knowledgeable, of high social status, accomplished, wealthy, and prosocial. The primary benefits leaders offer to followers are resolving conflicts, organizing cooperation, providing social functions, leading in inter-group conflicts, providing ritual services, and distributing resources. The primary benefits leaders receive from followers include material resources, increased social status, and social services from followers. The most frequent costs of leadership for leaders include loss of social status, resources or social services and increased risk of harm or conflict. For followers, the most frequent costs include loss of resources, costs in providing social services, and increased risk of harm or conflict.

Leadership in WEIRD settings often differs substantially from non-WEIRD settings and dominant theories in managerial studies have not previously had broad empirical data on leadership, management, and social organization across diverse populations. In WEIRD settings influence often extends into multiple domains, organizations are more hierarchical, and interactions between leaders and followers are less frequently face-to-face. A detailed view of leaders, their roles, and the costs and benefits of leadership across traditional societies and across contexts stands to make a significant empirical contribution for the development of more integrative evolutionary theories of leadership and social hierarchy.

@@

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limited generalizability of most extant psychological, sociological, and managerial data [@henrich_weirdest_2010], many social scientists are now diversifying their samples to better understand *human* psychology, behavior, organizations, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity among populations that are now often categorized as "non-WEIRD", "traditional", or "small-scale". At the same time, anthropologists have made the opposite mistake in often failing to recognize deep similarities among peoples of all cultures [@brown_human_1991; @said1979; @reyes-garcia_small-scale_2017].

Our results support both important dimensions of leadership that could be human universals, as well as systematic variation in other dimensions. We found leaders in the ethnographic record are portrayed as generally prosocial, are equipped with special competencies, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. Cross-cultural evidence also supports a dominance style of leadership involving coercive authority, punishment, aggression, and control of resources. These results generally provide broad cross-cultural support for several influential theories, such as the dominance-prestige model, the service-for-prestige model, the computational services model, and numerous collective action models [for review, see @garfield_evolutionary_2019].

Shamans emerged as an important category of leaders that are rarely discussed in the evolutionary literature [but see @singh_cultural_2017]. Supernatural qualities were among the top ten leader qualities across cultures (Figure \@ref(fig:varsupport)). Shamans share qualities with prestige-style leaders, such as having specialized knowledge and abilities, yet appear, at least at times, to use their knowledge to instill fear, similar to dominance-style leaders.

@@

Contemporary theoretical models of the evolution of leadership often emphasize leaders roles in organizing collective action and in managing between group conflicts [@hooper_theory_2010; @pietraszewski_evolution_2019; @price_evolution_2014]. Results here do support the important functions of leaders in organizing cooperation and facilitating social functions -- services which provide direct benefits to the group. However, the ubiquity of conflict resolution by leaders across social contexts and cultures calls for greater attention to the roles of leaders of all types in managing intra-group conflict between group members -- services tailored to provide direct benefits to a few individuals and more indirect group benefits [@glowacki_leadership_2015; @wiessner_collective_2019; @pupu_challenges_2018; @hagen_leadership_2019].

<!-- THINK WE ALREADY EXPRESSED THIS BELOW -->
The strong support for material and social benefits associated with leadership (Figure \@ref(fig:varsupportcb)) and the colinearity of measures associated with material wealth, social capital, and reproductive success (Figure \@ref(fig:cluster)A), suggests processes of leadership and followership might be strong drivers of the emergence and intergenerational transmission of inequality [@bowles_emergence_2010; @mattison_evolution_2016]. Various sources of material wealth and strategic or institutionalized processes of nepotism are widespread cross-culturally, including among many societies often described as "egalitarian." These results suggest anthropologists and economists investigating the emergence and evolution of inequality, including material, social, and reproductive dimensions, should incorporate and pay close attention to group processes and social strategies associated with leadership and followership.

@@

<!-- SUMMARY AND SUGGEST LEADERS BE PART OF OTHER GROUP RESEARCH -->
These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018]. Given the broad range of leadership dimensions represented in the ethnographic record, we suggest leaders-follower dynamics should be increasingly incorporated in evolutionary models behavior and group dynamics, including demography, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016].

>>>>

Leader-follower dynamics are central to much of social life. Processes of leadership and followership observed in nonindustrial societies and smaller-scale organizations are often distinct from those observed among WEIRD societies and their complex organizations. Many researchers have turned to evolutionary theory to integrate diverse perspectives on human leadership and to better understand the diversity of leadership and followership observed across human cultures and across social contexts. Evolutionary theories of leadership, in turn, draw on ethnographic case studies from traditional societies that are thought to more closely resemble the small, face-to-face societies in which humans evolved. Currently, though, there is limited systematic data on the nature of leadership in such societies.

We build on our previous cross-cultural findings to systematically evaluate dinebnsions of leadership across the full range of human cultural diversity. Over 1200 ethnographic texts from the Human Relations Area Files (HRAF) were coded on leader qualities, their functions, and the costs and benefits for both leaders and followers across diverse contexts. We discovered evidence for both cultural universals in leadership, as well as important variation with this sample of largely "non-WEIRD" populations. Leaders in the ethnographic record are most commonly described as intelligent and knowledgeable, of high social status, accomplished, wealthy, and prosocial. The primary benefits leaders offer to followers are resolving conflicts, organizing cooperation, providing social functions, leading in inter-group conflicts, providing ritual services, and distributing resources. The primary benefits leaders receive from followers include material resources, increased social status, and social services from followers. The most frequent costs of leadership for leaders include loss of social status, resources or social services and increased risk of harm or conflict. For followers, the most frequent costs include loss of resources, costs in providing social services, and increased risk of harm or conflict.

Leadership in WEIRD settings often differs substantially from non-WEIRD settings, where influence often extends into multiple domains, organizations are more hierarchical, and interactions between leaders and followers are less frequently face-to-face. The diversity within the non-WEIRD, however, should not be overlooked. Leadership theories, including evolutionary and managerial approaches have not previously had broad systematic, cross-cultural data on leadership, management, and social organization across diverse populations. A detailed view of leaders, their roles, and the costs and benefits of leadership across traditional societies and across contexts stands to make a significant empirical contribution for the development of more integrative evolutionary theories of leadership and social hierarchy.

@@

As Figure \@ref(fig:fivefold) reveals,

<!-- DAVIDS MARKET FORCES IDEAS -->
We agree with @pietraszewski_evolution_2019 that leadership and followership stem from biological market forces and information-processing roles [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019], however, much of human leadership goes beyond group coordination problems, such as leaders important roles in devising strategic outcomes between individuals, offering advice or decision-making services to individuals, and in making direct prosocial investments that do not involve substantial coordination. We therefore suggest leadership dynamics emerge from inter-individual competencies and leaders' abilities to provide benefits *or* impose costs, with these dimensions interacting with market forces rooted in follower preferences, group context, and cultural norms. 
 <!-- Whereas @pietraszewski_evolution_2019 emphasizes leaders ability to propose various enterprises and followers ability to evaluate them, -->

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- CAN'T JUST BE NONWEIRD -->
In recognizing the limited generalizability of most extant psychological, sociological, and managerial data [@henrich_weirdest_2010], many social scientists are now diversifying their samples to better understand *human* psychology, behavior, organizations, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity among populations that are now often categorized as "non-WEIRD", "traditional", or "small-scale". At the same time, anthropologists have made the opposite mistake in often failing to recognize deep similarities among peoples of all cultures [@brown_human_1991; @said1979; @reyes-garcia_small-scale_2017].

<!-- OUR SUPPORT FOR THEORIES -->
Our results support both important dimensions of leadership that could be human universals, as well as systematic variation in other dimensions. We found leaders in the ethnographic record are portrayed as generally prosocial, are equipped with special competencies, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. Cross-cultural evidence also supports a dominance style of leadership involving coercive authority, punishment, aggression, and control of resources. These results generally provide broad cross-cultural support for several influential theories, such as the dominance-prestige model, the service-for-prestige model, the computational services model, and numerous collective action models [for review, see @garfield_evolutionary_2019].

<!-- SHAMANS AND BEYOND DUAL MODEL -->
Shamans emerged as an important category of leaders that are rarely discussed in the evolutionary literature [but see @singh_cultural_2017]. Supernatural qualities were among the top ten leader qualities across cultures (Figure \@ref(fig:varsupport)). Shamans share qualities with prestige-style leaders, such as having specialized knowledge and abilities, yet appear, at least at times, to use their knowledge to instill fear, similar to dominance-style leaders.

@@

<!-- DIVERSE FUNCTIONS -->
Contemporary theoretical models of the evolution of leadership often emphasize leaders roles in organizing collective action and in managing between group conflicts [@hooper_theory_2010; @pietraszewski_evolution_2019; @price_evolution_2014]. Results here do support the important functions of leaders in organizing cooperation and facilitating social functions -- services which provide direct benefits to the group. However, the ubiquity of conflict resolution by leaders across social contexts and cultures calls for greater attention to the roles of leaders of all types in managing intra-group conflict between group members -- services tailored to provide direct benefits to a few individuals and more indirect group benefits [@glowacki_leadership_2015; @wiessner_collective_2019; @pupu_challenges_2018; @hagen_leadership_2019].

<!-- LEADER SUCCESS AND INEQUALITY -- REPETITIVE ? -->
The strong support for material and social benefits associated with leadership (Figure \@ref(fig:varsupportcb)) and the colinearity of measures associated with material wealth, social capital, and reproductive success (Figure \@ref(fig:cluster)A), suggests processes of leadership and followership might be strong drivers of the emergence and intergenerational transmission of inequality [@bowles_emergence_2010; @mattison_evolution_2016]. Various sources of material wealth and strategic or institutionalized processes of nepotism are widespread cross-culturally, including among many societies often described as "egalitarian." These results suggest anthropologists and economists investigating the emergence and evolution of inequality, including material, social, and reproductive dimensions, should incorporate and pay close attention to group processes and social strategies associated with leadership and followership.

@@

<!-- SUMMARY AND SUGGEST LEADERSHIP BE PART OF OTHER GROUP RESEARCH -->
These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018]. Given the broad range of leadership dimensions represented in the ethnographic record, we suggest leaders-follower dynamics should be increasingly incorporated in evolutionary models behavior and group dynamics, including demography, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016].
